CASE DETECTION OF TUBERCULOSIS PATIENTS USING MACHINE-LEARNING

### 2024-04-08 code/analysis-code/ML-analysis.qmd
Part4_1

>>>>

The evaluation of machine learning models' performance on test data is pivotal for gauging their efficacy and generalization capabilities. In this study, conducted to assess predictive accuracy, the models were measured using metrics such as accuracy and Area Under the Curve (AUC). The Lasso, Random Forest, and Support Vector models were compared against a Null Model, with their respective performance metrics outlined. The results revealed a notable performance gap between the advanced models and the Null Model. Specifically, the Random Forest model exhibited superior performance, achieving an accuracy of 81.57% and an AUC of 0.887. In contrast, the Lasso and Support Vector models attained accuracies of 80.26% and 79.68%, with corresponding AUC values of 0.889 and 0.882, respectively.

### 2024-04-08 code/eda-code/eda.qmd
Part4_1

# Create a summary table with specified variable types
summary_table_characteristics <- madaproject %>%
  select(all_of(variables), tb) %>%
  tbl_summary(
    by = tb,
    type = everything() ~ "categorical"
  )

# Print the summary table
summary_table_characteristics

# Save table
saveRDS(table1, file = here::here("results", "tables", "summary_table_characteristics .rds"))

>>>>

# Create a summary table with specified variable types
characteristics <- madaproject %>%
  select(all_of(variables), tb) %>%
  tbl_summary(
    by = tb,
    type = everything() ~ "categorical"
  )
characteristics <- data.frame(characteristics)
# Print the summary table
characteristics

# Save table

summarytable_file <- here("results", "tables", "characteristics.rds")
saveRDS(characteristics, file = summarytable_file)

### 2024-04-08 products/manuscript/Manuscript.qmd
Part4_1

---
title: "Prediction of Tuberculosis from Clinical Data"
subtitle: ""
author: Patrick Kaggwa
date: "2024-02-24"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/mada.bib
csl: ../../assets/apa.csl
---

@@

#Introduction Although early detection and treatment of Tuberculosis cases are the hallmark of successful TB control, diagnostic delay is still long and common in Uganda. Therefore, the study aimed atdeveloping a tool that uses machine-learning techniques to detect and diagnose TB more accurately in a shorter time. #Methods This was a retrospective study that used secondary data collected between 2011 to 2018 collected from health under Infectious diseases research collaboration an (IDRC) project called MIND (Mulago Inpatient Non-invasive Diagnosis for pneumonia). The study was conducted on data from 2296 tested patients of whom 1345 had no TB and 951 had TB. Different machine learning metrics(Accuracy, RMSE and AUC) were used evaluate the performance of the models. #Results {{< pagebreak >}}

@@

Tuberculosis (TB) remains a significant global health challenge due to diagnostic limitations affecting both pediatric and adult populations, resulting in delayed diagnosis or misdiagnosis[@Xin2019]. This delay impacts individual prognosis and community transmission[@f_verbeke_4_2012]. TB, caused by the bacterium Mycobacterium tuberculosis, primarily affects the lungs but can also involve other organs in its active form or remain latent without symptoms[@Bulled2020],. In 2020, TB ranked among the top 10 causes of death globally, with over 10 million new cases and 1.3 million deaths reported.

@@

How can TB prediction be modeled using early clinical patient data?

What is the performance of the TB prediction models?

@@

*Explain anything related to your statistical analyses.*

@@

```{r}
#| label: tbl-characteristics
#| tbl-cap: " Participants' Characteristics"
#| echo: FALSE
ch<- readRDS(here("results", "tables", "characteristics.rds"))

@@

##Determining requirements for Tuberculosis prediction that is appropriate data and modeling techniques

@@

```{r}
#| label: tbl-schematic2b
#| tbl-cap: "Model_perfomance."
#| echo: FALSE
mpf <- readRDS(here("results", "tables", "model_perfomance_train.rds"))

@@

```{r}
#| label: tbl-Random Forest
#| tbl-cap: " Model Predictions (Random Forest)"
#| echo: FALSE
rf <- readRDS(here("results", "tables", "rfconf_table.rds"))

@@

```{r}
#| label: fig-rf_ROCCURVE
#| fig-cap: "rf_ROCCURVE"
#| echo: FALSE
knitr::include_graphics(here("results","figures", "rf_ROCCURVE.png"))
```

```{r}
#| label: SVM
#| tbl-cap: "  Model Predictions (SVM)"
#| echo: FALSE
svm<- readRDS(here("results", "tables", "svmconf_table.rds"))
svm %>%
  kable()
```

@@

```{r}
#| label: fig-svm_ROCCURVE
#| fig-cap: "svm_ROCCURVE"
#| echo: FALSE
knitr::include_graphics(here("results","figures", "svm_ROCCURVE.png"))
```

```{r}
#| label: LASSO
#| tbl-cap: "  Model Predictions (LASSO)"
#| echo: FALSE
ls<- readRDS(here("results", "tables", "lassoconf_table.rds"))
ls %>%
  kable()
```

@@

```{r}
#| label: fig-Lasso_ROCCURVE
#| fig-cap: "Lasso_ROCCURVE"
#| echo: FALSE 
knitr::include_graphics(here("results","figures", "Lasso_ROCCURVE.png"))
```
```{r}
#Figure1. Random Forest ROC Curve

```

#Models Evaluation {{< pagebreak >}}

@@

The aim of the study was to apply machine-learning techniques in developing models for TB screening and detection in a shorter time based on patient data records to contribute to disease notification. The application of machine learning approaches have therefore been found to be successful in prediction [@Holzinger].

## Summary and Interpretation

## Strengths and Limitations

## Conclusions

>>>>

---
title: "CASE DETECTION OF TUBERCULOSIS PATIENTS USING MACHINE-LEARNING"
subtitle: ""
author: Patrick Kaggwa
date: "2024-02-24"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/mada.bib
csl: ../../assets/apa.csl
---

@@

## Introduction

Although early detection and treatment of Tuberculosis cases are the hallmark of successful TB control, diagnostic delay is still long and common in Uganda. Therefore, the study aimed atdeveloping a tool that uses machine-learning techniques to detect and diagnose TB more accurately in a shorter time.

## Methods

This was a retrospective study that used secondary data collected between 2011 to 2018 collected from health under Infectious diseases research collaboration an (IDRC) project called MIND (Mulago Inpatient Non-invasive Diagnosis for pneumonia). The study was conducted on data from 2296 tested patients of whom 1345 had no TB and 951 had TB. Different machine learning metrics(Accuracy, RMSE and AUC) were used evaluate the performance of the models.

## Results
Evaluation of four machine learning models—LASSO, Random Forest, Support Vector Machine (SVM), and Null model—revealed notable performance differences. Notably, the Random Forest model achieved the highest accuracy at 81.57% and an AUC of 0.8868, outperforming the other models.

## Consclusion

These results highlight the potential of advanced machine learning algorithms, particularly Random Forest, in enhancing TB diagnosis and early detection strategies, offering valuable insights for clinical decision-making and the development of more effective screening protocols.
{{< pagebreak >}}

@@

Tuberculosis (TB) remains a significant global health challenge due to diagnostic limitations affecting both pediatric and adult populations, resulting in delayed diagnosis or misdiagnosis[@Xin2019]. This delay impacts individual prognosis and community transmission[@]. TB, caused by the bacterium Mycobacterium tuberculosis, primarily affects the lungs but can also involve other organs in its active form or remain latent without symptoms[@Bulled2020],. In 2020, TB ranked among the top 10 causes of death globally, with over 10 million new cases and 1.3 million deaths reported.

@@

1.  How can TB prediction be modeled using early clinical patient data?

2.  What is the performance of the TB prediction models?

@@

@@

```{r }
#| label: tbl-characteristics
#| tbl-cap: " Participants' Characteristics"
#| echo: FALSE
ch<- readRDS(here("results", "tables", "characteristics.rds"))

@@

## Determining requirements for Tuberculosis prediction that is appropriate data and modeling techniques

@@

```{r }
#| label: tbl-model_perfomance_train
#| tbl-cap: "Model_perfomance."
#| echo: FALSE
mpf <- readRDS(here("results", "tables", "model_perfomance_train.rds"))

@@

```{r }
#| label: tbl-Random Forest
#| tbl-cap: " Model Predictions (Random Forest)"
#| echo: FALSE
rf <- readRDS(here("results", "tables", "rfconf_table.rds"))

@@

```{r }
#| label: fig-Random Forest_Roccurve
#| fig-cap: "Random Forest_Roccurve"
#| echo: FALSE
knitr::include_graphics(here("results","figures", "rf_ROCCURVE.png"))
```

```{r }
#| label: tbl-Support Vector Machine
#| tbl-cap: "  Model Predictions (SVM)"
#| echo: FALSE
svm<- readRDS(here("results", "tables", "svmconf_table.rds"))
svm %>%
  kable()
```

@@

```{r }
#| label: fig-supporter Vector Machine_Roccurve
#| fig-cap: "svm_ROCCURVE"
#| echo: FALSE
knitr::include_graphics(here("results","figures", "svm_ROCCURVE.png"))
```

```{r }
#| label: tbl-LASSO
#| tbl-cap: "  Model Predictions (LASSO)"
#| echo: FALSE
ls<- readRDS(here("results", "tables", "lassoconf_table.rds"))
ls %>%
  kable()
```

@@

```{r }
#| label: fig-Lasso_ROCCURVE
#| fig-cap: "Lasso_ROCCURVE"
#| echo: FALSE 
knitr::include_graphics(here("results","figures", "Lasso_ROCCURVE.png"))
```

```{r}
#| label: model_perfomance_test
#| tbl-cap: "model_perfomance_test"
#| echo: FALSE
tmpf <- readRDS(here("results", "tables", "model_perfomance_test.rds"))

tmpf %>%
  kable()
```

# Models Evaluation

The evaluation of machine learning models' performance on test data is pivotal for gauging their efficacy and generalization capabilities. The predictive accuracy of the models were measured using the test data. The Lasso, Random Forest, and Support Vector models were compared against a Null Model, with their respective performance metrics outlined. The results revealed a notable performance gap between the advanced models and the Null Model. Specifically, the Random Forest model exhibited superior performance, achieving an accuracy of 81.57% and an AUC of 0.887. In contrast, the Lasso and Support Vector models attained accuracies of 80.26% and 79.68%, with corresponding AUC values of 0.889 and 0.882, respectively.

@@

The study presents a comprehensive analysis of machine learning models' efficacy in predicting the presence of tuberculosis (TB) based on a dataset comprising 2296 instances and 14 variables associated with TB. The study selected nine variables for prediction after conducting chi-square tests to eliminate non-statistically significant variables associated with the outcome. The retained variables for prediction included asthma, alcohol use, fever, chest radiography, home fuel use, weight loss, presence of blood in sputum, oxygen saturation, and HIV status. The study evaluated four distinct machine learning models, namely LASSO, Random Forest, Support Vector Machine (SVM), and a Null model, for their predictive performance using key metrics accuracy and Area Under the Curve (AUC). The LASSO model achieved an accuracy of 80.26% and an AUC of 0.8895. The Random Forest model outperformed others with an accuracy of 81.57% and an AUC of 0.8868. SVM achieved an accuracy of 79.68% and an AUC of 0.8825. The Null model achieved an accuracy of 60.23% and an AUC of 0.5.

Furthermore, the study presented detailed classification results for each model, emphasizing their ability to correctly classify true positives and true negatives while minimizing false positives and false negatives. Particularly, the Random Forest model demonstrated robust performance in correctly identifying true positives and true negatives, with minimal misclassifications. The findings underscore the potential of advanced machine learning algorithms, particularly Random Forest and LASSO, in enhancing TB diagnosis and early detection strategies. The manuscript provides valuable insights into the predictive capabilities of different machine learning models which can inform clinical decision-making and aid in the development of more effective TB screening tolls and protocols.

# Strengths

In this study, a rigorous variable selection process was implemented, utilizing chi-square tests to identify statistically significant predictors associated with tuberculosis (TB) from a dataset comprising 2296 instances and 14 variables. By employing stringent selection criteria, the study ensures that only variables with meaningful associations with TB are included in subsequent analyses, thereby enhancing the model's interpretability, and reducing the risk of overfitting. Moreover, the evaluation of multiple machine learning models, including LASSO, Random Forest, Support Vector Machine (SVM), and a Null model, contributes to a comprehensive comparison of different predictive approaches. This assessment allows for informed decision-making regarding model selection, facilitating the identification of the most suitable algorithm for TB prediction.

Additionally, the study employs robust performance metrics such as accuracy and Area Under the Curve (AUC) to evaluate model performance objectively. These metrics provide a comprehensive assessment of each model's predictive capabilities, enabling comparisons across different algorithms. Furthermore, the detailed classification results presented offer insights into the models' ability to correctly classify true positives and true negatives while minimizing false positives and false negatives.

# Limitations and Recommendations

Despite the strengths, this study had limitations that warrant consideration. Firstly, the dataset's potential class imbalance, characterized by variations in the number of instances for individuals with and without TB, may impact model performance of the model. Additionally, while the variable selection methodology employed chi-square tests, alternative feature selection methods like recursive feature elimination could offer further insights into the predictive variables' relevance and importance. Future research should aim incorporate larger, more diverse datasets and conducting external validation studies to assess model performance in real-world clinical settings and use of more robust techniques such as neural networks.

# Conclusions

The findings underscore the potential of machine learning algorithms, particularly Random Forest and LASSO, in enhancing TB diagnosis and early detection strategies. Future research should further advance the development of effective TB screening protocols and contribute to improved TB diagnosis and management on a global scale.
